Comparing corpora
-----------------

It can be interesting to compare vocabulary use in two corpora. For instance, to find out:
- which terms certain actors are more/less likely to use compared to other actors.
- which terms are more/less common in texts about certain actors/issues compared to other actors/issues.
- whether the use of certain terms has increased/decreased between two periods in time.

In this howto we demonstrate how to compare vocabulary use in two corpora, and offer a function to do so conveniently. 
For data we use wikinews articles about Iraq.

```{r}
library(corpustools)
load('wikinews_iraq_dtm_meta.rdata')
dtm
head(meta)
```

For example, lets split our corpus into those articles that mention Bush and those that do not:

```{r}
ncol(dtm)
w = as.matrix(dtm[,"Bush"])
dtm.bush = dtm[w>0, ]
dtm.rest = dtm[w==0, ]
```

Now, we can compute and compare the term frequencies for both. First, we use the `term.statistics` function to get the term statistics (including term frequency) for both corpora. 

```{r}
terms.bush = term.statistics(dtm.bush)
terms.rest = term.statistics(dtm.rest)
head(terms.bush)
```

We then match the term frequencies of dtm.rest to the term statistics of dtm.bush.
```{r}
freqs.rest = terms.rest[, c("term", "termfreq")]
terms.bush = merge(terms.bush, freqs.rest, all.x=TRUE, by="term")
terms.bush[is.na(terms.bush)] = 0
head(terms.bush)
```

`terms.bush` now contains the term statistics in the articles mentioning bush, 
as well as the frequency in the reference corpus consisting of the other articles.
We can now compute and sort by the overrepresentation of terms:

```{r}
terms.bush$relfreq.x = terms.bush$termfreq.x / sum(terms.bush$termfreq.x)
terms.bush$relfreq.y = terms.bush$termfreq.y / sum(terms.rest$termfreq)
terms.bush$over = terms.bush$relfreq.x / (terms.bush$relfreq.y + .001)
terms.bush = terms.bush[order(-terms.bush$over), ]
head(terms.bush, n=10)
```

This gives a list of the words that occur 'too much' in the articles mentioning Bush,
or in other words the collocates of the word 'Bush'.
To make this easier, and to also provide statistical association measures such as chi-squared,
the function `corpora.compare` is provided. 
The following example selects all words that are underrepresented in the `bush` corpus,
and sorts them by chi-squared:

```{r}
terms = corpora.compare(dtm.bush, dtm.rest)
terms = terms[terms$over < 1, ]
terms = terms[order(-terms$chi),]
head(terms)
```

What can be seen from these two word lists is that the articles mentioning Bush are more political in nature,
while the other articles describe more (military) action. 

Of course, this can also be used to compare e.g. vocabulary differences between newspapers, speakers, periods, etc.
For example, the following uses the article metadata to compare vocabulary after 2012 with the vocabulary before that date.

```{r message=FALSE}
dtm.before = dtm[meta$date < as.Date('2012-01-01'),]
dtm.after = dtm[meta$date >= as.Date('2012-01-01'),]
terms = corpora.compare(dtm.after, dtm.before)
terms = terms[order(-terms$chi),]
head(terms[terms$over > 1, ])
head(terms[terms$over < 1, ])
```

So, the later articles mention the scandals with Abu Ghraib and phosphorus munition more frequently, 
while the earlier articles metion the constitution and refugees. 